---
title: "GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Functional class sorting is widely used for pathway analysis using tools like GSEA,
yet there is no consensus in how this can be conducted for Illumina Infinium
methylation array data.

Here we propose a simple approach which involves the following:

1. Limma test on probes.

2. For each gene, calculate the median t-statistic from step 1.

3. Use this median t-stat in Camera pre-ranked test for gene sets.

In this example, I'm using matched infinium EPIC 850k data from (n=37) normal and lung cancer samples (GSE158422).
The data was previously preprocessed and normalised using the minfi package (see the folder called "misc").

Here the gene sets are obtained from Reactome.

## Requirements

This analysis was run on a 8C/16T computer with 64GB RAM running at 3.8 GHz.
This workflow used 34 GB RAM and took
RAM usage can be moderated by reducing the parallel core count.

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("kableExtra")
  library("beeswarm")
})

```

## Load data

* annotations

* probe sets

* gene sets

* design matrix

* mval matrix

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])

gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
sets <- split(rep(names(gp2), lengths(gp2)), unlist(gp2))

summary(unlist(lapply(sets,length)))

gmt_import <- function(gmtfile) {
    genesetLines <- strsplit(readLines(gmtfile), "\t")
    genesets <- lapply(genesetLines, utils::tail, -2)
    names(genesets) <- unlist(lapply(genesetLines, head, 1))
    attributes(genesets)$originfile <- gmtfile
    if( any(duplicated(names(genesets))) ) {
        warning("Duplicated gene sets names detected")
    }
    genesets
}

genesets <- gmt_import("https://ziemann-lab.net/public/gmea_prototype/ReactomePathways.gmt")

if (!file.exists("GSE158422_design.rds")) {
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_design.rds", "GSE158422_design.rds")
}
design <- readRDS("GSE158422_design.rds")

if (!file.exists("GSE158422_design.rds")) {
 download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
}
mval <- readRDS("GSE158422_mx.rds")

```

## Gene level analysis

Performs limma on probes and then reports the median limma t-statistic for each gene.

It also applies a one-sample t-test for probes belonging to each gene, but that isn't
used in downstream gene set analysis.

```{r,pmea}

pmea <- function(mval,design,sets,cores=2) {
  fit <- lmFit(mval, design)
  fit <- eBayes(fit)
  top <- topTable(fit,coef=ncol(design),num=Inf, sort.by = "P")
  l <- mclapply(seq(1,length(sets)), function(i) {
    g <- names(sets[i])
    tstats <- top[rownames(top) %in% sets[[i]],"t"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    if ( length(tstats) < 2 ) {
      pval=1
    } else {
      wtselfcont <- t.test(tstats)
      pval=wtselfcont$p.value
    }
    c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "P.Value"=pval)
  } , mc.cores=cores)
  df <- do.call(rbind, l)
  rownames(df) <- df[,1]
  df <- df[,-1]
  tmp <- apply(df,2,as.numeric)
  rownames(tmp) <- rownames(df)
  df <- as.data.frame(tmp)
  df$sig <- -log10(df[,4])
  df <- df[order(-df$sig),]
  df$FDR <- p.adjust(df$P.Value)
  out <- list("df"=df,"toptable"=top)
  return(out)
}

tic()
res <- pmea(mval,design,sets,cores=8)
resdf <- res$df
toc()

head(resdf,20) %>%
  kbl(caption="GMEA gene level analysis top results (p-value)") %>%
  kable_styling(full_width=FALSE)

```

## Gene set analysis

Now perform gene set analysis using CameraPR with reactome gene sets.

```{r,camerapr_gene_set}

tic()
stat <- resdf$median
names(stat) <- rownames(resdf)
stat[is.na(stat)] <- 0
cres <- cameraPR(statistic=stat, index=genesets, use.ranks = FALSE, inter.gene.cor=0.01, sort = FALSE)
cres$ES <- unlist(lapply(genesets, function(gset) {median(stat[which(names(stat) %in% gset )]) }))
cres <- cres[order(cres$PValue),]
toc()

top <- head(cres,10)

top %>%
  kbl(caption="CameraPR gene set results") %>%
  kable_styling(full_width=FALSE)

top <- top[order(top$ES),]
topsets <- genesets[names(genesets) %in% rownames(top)]

scores <- lapply(topsets, function(gset) { stat[which(names(stat) %in% gset )] })
scores <- scores[order(unlist(lapply(scores,median)))]

par(mar=c(5,25,3,1))
boxplot(scores,horizontal=TRUE,las=1,cex=0,cex.axis=0.9,xlab="Enrichment score")
mtext("Top pathways in lung cancer")
beeswarm(scores,add=TRUE,horiz=TRUE,pch=19,cex=0.3)
grid()

```

## Session information

```{r,save}

sessionInfo()

```
